Modelling the Lymphatic Metastatic Progression Pathways of OPSCC from Multi-Institutional Datasets

### 2023-01-06 src/tex/ms.tex
[showyourwork] overleaf sync

\begin{table*}
    \centering
    \input{output/data_table.tex}
    \caption{Prevalence of involvement patterns in the multi-centric dataset. An involvement pattern is characterized by the state of the six \glspl{lnl}: A red dot means the \gls{lnl} was reported to be metastatic, a green dot means it was determined to be healthy and a question mark means that the prevalence was marginalized over the state of this \gls{lnl}.}
    \label{table:data_prevalence}
\end{table*}

@@

This work concerns itself with \gls{opscc}, which commonly spreads lymphatically. The only two publicly available datasets on nodal involvement patterns in \gls{opscc}, at the time of writing, show that 80\% of patients where either cN+ or -- if available -- pN+ \cite{ludwig_dataset_2022}. Since the exact number and location of lymph nodes varies from patient to patient, the nodes are grouped into anatomically defined \glspl{lnl}. These levels are then often irradiated or resected prophylactically due to the risk of harboring occult metastases despite negative findings from imaging modalities. In current clinical practice, \gls{ctv-n} definition is instead mostly based on guidelines \cite{gregoire_ct-based_2003,gregoire_delineation_2014,gregoire_delineation_2018,eisbruch_intensity-modulated_2002,biau_selection_2019,chao_determination_2002,vorwerk_guidelines_2011,ferlito_elective_2009} that are derived from the observed prevalence of involvement in an \gls{lnl} for a given tumor location.

@@

A model for estimating the risk of microscopic disease, given a personal diagnosis based on \glspl{bn} was previously developed \cite{pouymayou_bayesian_2019} and subsequently adapted to make use of \glspl{hmm} to include T-category in an intuitive manner \cite{ludwig_hidden_2021}. These models were trained on a dataset of early T-category (T1 and T2) patients that could be reconstructed from \cite{sanguineti_defining_2009}. Since this publication only includes details about the involvement of the ipsilateral \glspl{lnl} I, II, III, and IV, the models were limited to these nodal levels as well.

@@

\begin{figure}
    \script{thermo_int.py}
    \centering
    \includegraphics[width=\linewidth]{figures/thermo_int.png}
    \caption{Placeholder}
    \label{fig:thermo_int}
\end{figure}

\begin{table*}
    \centering
    \begin{tabular}{ | c | c | c | }
        \hline
        $K_\text{1v2}$ & $\ln{K_\text{1v2}}$ & support for $\mathcal{M}_1$ \\
        \hline
        $< 10^0$ & $< 0$ & negative evidence (supports $\mathcal{M}_2$) \\
        $10^0$ to $10^{\nicefrac{1}{2}}$ & 0 to 1.15 & barely worth a mention \\
        $10^{\nicefrac{1}{2}}$ to $10^1$ & 1.15 to 2.3 & substantial \\
        $10^1$ to $10^{\nicefrac{3}{2}}$ & 2.3 to 3.45 & strong \\
        $10^{\nicefrac{3}{2}}$ to $10^2$ & 3.45 to 4.6 & very strong \\
        $> 10^2$ & $> 4.6$ & decisive \\
        \hline
    \end{tabular}
    \caption{Interpretation of Bayes factors and their natural logarithms in terms of their support for or against one of the two compared models as introduced by Harold Jeffreys \cite{jeffreys_theory_1998}.}
    \label{table:bayes_factor}
\end{table*}

@@

@@

@@

\begin{table*}
    \centering
    \input{output/results_table.tex}
    \caption{Model comparison results from the base graph and the extended graph we chose as the ``winnning'' model. For both \glspl{dag} we show the log-evidence, computed via thermodynamic integration, the negative one half of the \gls{bic}, as well as the maximum log-likelihood that was encountered during the final \gls{mcmc} sampling round.}
    \label{table:evidence}
\end{table*}

@@

@@

\begin{figure}
    \script{comp_IIandVII_prevs.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/comp_IIandVII_prevs.png}
        \caption{Placeholder.}
        \label{fig:IIandVII_prevs}
    \end{centering}
\end{figure}

\begin{figure}
    \script{comp_IIIandV_prevs.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/comp_IIIandV_prevs.png}
        \caption{Placeholder.}
        \label{fig:IIIandV_prevs}
    \end{centering}
\end{figure}

\begin{figure}
    \script{comp_IVandV_prevs.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/comp_IVandV_prevs.png}
        \caption{Placeholder.}
        \label{fig:IVandV_prevs}
    \end{centering}
\end{figure}

>>>>

@@

This work concerns itself with \gls{opscc}, which commonly spreads lymphatically. The only two publicly available datasets on nodal involvement patterns in \gls{opscc}, at the time of writing, show that 80\% of patients where either cN+ or -- if available -- pN+ \cite{ludwig_dataset_2022}. Since the exact number and location of lymph nodes varies from patient to patient, the nodes are grouped into anatomically defined \glspl{lnl}. These levels are then often irradiated or resected prophylactically due to the risk of harboring occult metastases despite negative findings from imaging modalities. In current clinical practice, \gls{ctv-n} definition is mostly based on guidelines \cite{gregoire_ct-based_2003,gregoire_delineation_2014,gregoire_delineation_2018,eisbruch_intensity-modulated_2002,biau_selection_2019,chao_determination_2002,vorwerk_guidelines_2011,ferlito_elective_2009} that are derived from the observed prevalence of involvement in an \gls{lnl} for a given tumor location.

@@

A model for estimating the risk of microscopic disease, given a personal diagnosis based on \glspl{bn} was previously developed \cite{pouymayou_bayesian_2019}. It was subsequently formulated as a \glspl{hmm} to include T-category in an intuitive manner \cite{ludwig_hidden_2021}. These models were trained on a dataset of early T-category (T1 and T2) patients that could be reconstructed from \cite{sanguineti_defining_2009}. Since this publication only includes details about the involvement of the ipsilateral \glspl{lnl} I, II, III, and IV, the models were limited to these nodal levels as well.

@@

\begin{figure}
    \script{thermo_int.py}
    \centering
    \includegraphics[width=\linewidth]{figures/thermo_int.png}
    \caption{The expected log-accuracy under the power posterior plotted against the value of the inverse temperature $\beta$. These values were computed during a \gls{ti} run for the base graph model (red line), as well as the winning graph model (green line). Note that the scaling of the x-axis is chosen such that the 64 points on the temperature ladder appear evenly spaced. This is to stress how the log-accuracy develops in the range from $\beta=0$ to around $\beta=0.1$. The area under these two curves yields an approximation of the log-evidence of the respective model. In this plot one can see that due to the simplicity of the base graph model its accuracy increases earlier and faster that the winning graph model's. However, already around $\beta=0.1$ the model based on the winning graph outperforms the base graph model. Ultimately, the winning graph's log-evidence is better than the base graph model's by a value of around 3 (see \cref{table:evidence}).}
    \label{fig:thermo_int}
\end{figure}

\begin{table}
    \centering
    \begin{tabular}{ | c | c | c | }
        \hline
        $K_\text{1v2}$ & $\ln{K_\text{1v2}}$ & support for $\mathcal{M}_1$ \\
        \hline
        $< 10^0$ & $< 0$ & negative (supports $\mathcal{M}_2$) \\
        $10^0$ to $10^{\nicefrac{1}{2}}$ & 0 to 1.15 & barely worth a mention \\
        $10^{\nicefrac{1}{2}}$ to $10^1$ & 1.15 to 2.3 & substantial \\
        $10^1$ to $10^{\nicefrac{3}{2}}$ & 2.3 to 3.45 & strong \\
        $10^{\nicefrac{3}{2}}$ to $10^2$ & 3.45 to 4.6 & very strong \\
        $> 10^2$ & $> 4.6$ & decisive \\
        \hline
    \end{tabular}
    \caption{Interpretation of Bayes factors and their natural logarithms in terms of their support for or against one of the two compared models as introduced by Harold Jeffreys \cite{jeffreys_theory_1998}.}
    \label{table:bayes_factor}
\end{table}

@@

\subsection{Investigating Spread Graphs}
\label{subsec:methods:graphs}

The \gls{dag} shown in \cref{fig:graph_with_obs} includes the \glspl{lnl} I, II, III, and IV since previous models \cite{pouymayou_bayesian_2019,ludwig_hidden_2021} had to rely on a reconstructed dataset of \gls{opscc} patients that contained only information on these four levels \cite{sanguineti_defining_2009}. Now, with datasets available that almost always report the \glspl{lnl} I, II, III, IV, V, and VII, we may add more \glspl{rv} to extend the graph and investigate how these \glspl{rv} should be connected among each other.

First, we notice that the direct arcs from tumor to each of the \glspl{lnl} must be present, since every \gls{lnl} appears metastatic in isolation at least once. For example, some patients presented with metastases in \gls{lnl} I, while the other levels appeared healthy. If no spread was allowed from the tumor to $X_1$ (i.e., $b_i = 0$), the likelihood of observing this patient would be zero.

We have more freedom in choosing how to connect the \glspl{lnl} to each other.

@@

\begin{figure}
    \centering
    \includegraphics{figures/extended_graph.png}
    \caption{Extended \gls{dag} representing different possible spread graphs underlying the \gls{hmm}. As in \cref{fig:graph_with_obs}, red arcs are parametrized with probabilities of spread from the tumor (red circle) to the \glspl{lnl} (blue circles). These red arcs, together with the blue arcs fron \gls{lnl} to \gls{lnl}, make up the \emph{base graph}. One after the other, each of the green arcs was added to the base graph. Subsequently, the performance of the resulting models in terms of its \gls{bic} was compared to the base graph to asssess whether the additional edge should be kept in the \emph{winning graph} or not.}
    \label{fig:extended_graph}
\end{figure}

\begin{table}
    \centering
    \variable{output/data_table.tex}
    \caption{Prevalence of involvement patterns in the multi-centric dataset. An involvement pattern is characterized by the state of the six \glspl{lnl}: A red dot means the \gls{lnl} was reported to be metastatic, a green dot means it was determined to be healthy and a question mark means that the prevalence was marginalized over the state of this \gls{lnl}.}
    \label{table:data_prevalence}
\end{table}

@@

\begin{table}
    \centering
    \variable{output/results_table.tex}
    \caption{Model comparison results from the base graph and the extended graph we chose as the ``winnning'' model. For both \glspl{dag} we show the log-evidence, computed via thermodynamic integration, the negative one half of the \gls{bic}, as well as the maximum log-likelihood that was encountered during the final \gls{mcmc} sampling round.}
    \label{table:evidence}
\end{table}

@@

In \cref{fig:IandII_low_prevs,fig:IandII_high_prevs,fig:IIandVII_prevs,fig:IIIandV_prevs,fig:IVandV_prevs} the base graph's predictions for involvement patterns that include the \glspl{lnl} I, V, and VII have been plotted as gray, hatched histograms alongside the corresponding prevalence in the data and the prediction by the winning graph. This allows us to compare how the winning graph improves its fit to the data in these particular scenarios around low-risk \glspl{lnl}. How frequently each of the investigated lymphatic metastatic patterns were observed in the data is additionally tabulated in \cref{table:data_prevalence}, both for early and for late T-category cases.

@@

\begin{figure}
    \script{comp_IIandVII_prevs.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/comp_IIandVII_prevs.png}
        \caption{Prevalences for involvement patterns where \gls{lnl} VII is involved with (red) and without (blue) \gls{lnl} II. The solid lines depict a Beta posterior over the prevalence as observed in the data. The colored histograms show these prevalences as predicted by the winning graph. The hatched histograms show the same predictions from the base graph for comparison. Since the winning graph does not include an additional arc between \glspl{lnl} II and VII, the fit w.r.t. these two \glspl{LNL} does not improve.}
        \label{fig:IIandVII_prevs}
    \end{centering}
\end{figure}

\begin{figure}
    \script{comp_IIIandV_prevs.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/comp_IIIandV_prevs.png}
        \caption{Observations of prevalences in the data (solid lines) compared to predictions from the base (hatched histograms) and the winning graph model (colored histograms). The investigated involvement patterns here are how often is \gls{lnl} V involved together with \gls{lnl} III (red) and how often is level V involved while \gls{lnl} III is healthy (green). For early T-category cases, the fit of both models to the data is good. For late T-category, however, the winning graph shows a better ability to split the two investigated patterns of involvement. This is presumably due to the added connection from \gls{lnl} III to V.}
        \label{fig:IIIandV_prevs}
    \end{centering}
\end{figure}

\begin{figure}
    \script{comp_IVandV_prevs.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/comp_IVandV_prevs.png}
        \caption{Predictions vs observations of the prevalence of two patterns of involvement: One for the case when both \gls{lnl} VI and V are metastatic (blue) and one for the case where \gls{lnl} V is involved with level IV being healthy (orange). Again, the solid lies depict the Beta posterior over the prevalence of these patterns in the data. Colored histograms show the predicted distribution over the two patterns from the winning graph model. Hatched histograms show the same predictions from the base graph model for comparison. Although no arc was added from \gls{lnl} IV to V, the winning graph slightly improves its predictions both for early (top panel) and late T-category cases (bottom panel).}
        \label{fig:IVandV_prevs}
    \end{centering}
\end{figure}
